---
title: "Lab 06 - Ugly charts and Simpson's paradox"
author: "Lily Botha"
date: "02/23/2026"
output: github_document
---

### Load packages and data

```{r load-packages, message=FALSE}
library(tidyverse) 
library(dsbox)
library(mosaicData) 
library(dplyr)
library(ggplot2)
library(countrycode)
```

### Exercise 1

```{r read-instructional-staff-data, message=FALSE}
staff <- read_csv("data/instructional-staff.csv")
```

```{r long-format, message=FALSE}
staff_long <- staff %>%
  pivot_longer(cols = -faculty_type, names_to = "year") %>%
  mutate(value = as.numeric(value))
staff_long
```

I improved this graph by using a clean line plot with clear labels, a descriptive title, and a legend. 

```{r staff-line-plot}
staff_long %>%
  ggplot(aes(
    x = year,
    y = value,
    group = faculty_type,
    color = faculty_type
  )) +
  labs(
    title = "Trends in Instructional Staff Employment Status",
    x = "Year",
    y = "Percent of Total Instructional Staff",
    color = "Faculty Type"
  ) +
  geom_line(linewidth = 1)
```

### Exercise 2

You could make part-time the focus by making it one bold color, and making the other lines grey. Or, you could directly label the part-time trend line. 

### Exercise 3

```{r read-fisheries-data}
fisheries <- read_csv("data/fisheries.csv")
```
Issues: The pie charts are very unhelpful. There are way too many slices, so its impossible to make any meaningful comparisons. The chart on the left is also confusing. You can't see all the countries on the x-axis, there's no clear grouping, and it doesn't really match with the pie charts. There is no clear takeaway from these plots. 

Improvements: I want to get rid of the pie charts and instead use sorted bar charts that are grouped. Instead of using tons, I want to convert to percent of production for easier comparisons. The countries could be sorted by capture percentage, and the plots should be faceted by continent to reduce clutter. I can show the aquaculture percentage by using color fill, and I will also add better labels/themes. 

Aspirational improvements: Maybe it would be helpful to show only the top producing countries per continent. There are still too many countries per facet, so the plot looks dense. 

```{r fisheries-plot}
fisheries %>%
  group_by(country) %>%
  summarise(
    capture = sum(capture, na.rm = TRUE),
    aquaculture = sum(aquaculture, na.rm = TRUE),
    total = sum(total, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    percentcapture = 100 * capture / total,
    percentaquaculture = 100 * aquaculture / total,
    continent = suppressWarnings(countrycode(country,
      origin = "country.name",
      destination = "continent"
    )),
    continent = if_else(is.na(continent), "Other/Unknown", continent),
    continent = factor(continent),
    country = reorder(country, percentcapture)
  ) %>%
  ggplot(aes(x = country, y = percentcapture, fill = percentaquaculture)) +
  geom_col() +
  coord_flip() +
  facet_wrap(~ continent, scales = "free_y") +
  labs(
    x = "",
    y = "Percent of production from capture",
    title = "Fishery Production by Country",
    subtitle = "Bar height = % capture, color = % aquaculture",
    fill = "% aquaculture"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    axis.text.y = element_text(size = 4)
  )
```
```


















